A note on observation processes in epidemic models

### 2019-07-23 ms/ms.tex
very close

\title{A note on observational processes in epidemic models}
\author{Sang Woo Park and Benjamin M. Bolker}

@@

@@

Mechanistic analyses of epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an epidemic \citep{breto2009time}. In order to make a 
precise and accurate inference, disease modelers often focus on developing
more realistic process models. For example, the same London measles time 
series from the prevaccination era has been analyzed using multiple models -- 
these models account for time-varying transmission rates \citep{fine1982measles}, 
realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

Despite the amount of effort put into developing \emph{better} process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Most disease models assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or develops symptom; one exception is
the model by \citep{he2009plug}, which assumes that infections are counted upon
recovery (because diagnosed cases are put to bed rest and are effectively 
no longer infectious).

Here, we use a simple Susceptible-Infected-Removed (SIR) model to study how 
assumptions about the underlying observation process affect parameter estimates
of the SIR model. We show that making wrong assumptions about the timing of 
incidence reports can lead to biases in parameter estimates and narrow 
confidence intervals.

@@

The Susceptible-Infected-Removed (SIR) model describes how disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate, $\gamma$ is the recovery rate, and $N = S + I + R$ is
the total population size. We define incidence at time $t$ as the number of newly infected
individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
the reporting time step. We expect infected individuals to report their infection some time
after their infection; the number of reported cases defines the \emph{observed} incidence.

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection ($i_1(t)$) can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt,
\end{equation}
where $\Delta t$ is the reporting time step.
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model under-reporting using a negative binomial distribution with
mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
likewise, we will refer to epidemic time series generated from these models 
as infection time series and recovery time series.

@@

In order to introduce the problem, we first compare the dynamics of observed 
incidence, $i_1(t)$ and $i_2(t)$, (Figure 1). Lags in reporting time delay 
the observed epidemic peak timing and reduce the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
would not expect to be able to distinguish between the two reporting processes.

Note that incidence measured at these two periods are different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the dynamics of observed incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While it is
uncommon, some models do not make a clear distinction between the two.

\begin{figure}
\includegraphics[width=\textwidth]{../figure/example.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
We used the following parameters to simulate the SIR model and generate the figure: 
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
and $\Delta t = 0.1$ time units.
}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../figure/compare_deterministic.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
We used the following parameters to simulate the SIR model: 
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
$\theta = 10$, and $\Delta t = 0.1$ time units.
}
\end{figure}

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (or the 
transmission rate). When we assume that the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in clearer bias and lower coverage (Figure 2B); however, 
bias from fitting the wrong model changes direction (e.g., fitting the infection
model to recovery model overestimates the reproduction number instead).

Differences in the direction of the bias can be explained by the estimates of the
exponential growth rate ($r = \beta - \gamma$) and the mean infectious period ($1/\gamma$).
When we fit the recovery model to infection time series, the model overestimates
the initial growth rate in order to match the higher epidemic peak of the 
infection time series. When the mean infectious period is fixed, higher growth rate
translates to higher basic reproduction number as we see in Figure 2B. When we allow
the mean infectious period to vary, the model underestimates the mean infectious
period (high $\gamma$), which also decreases the estimate of the basic reproduction number.
Likewise, fitting the infection model to recovery time series overestimates the
mean infectious period (low $\gamma$), increasing the estimate of the basic reproduction number;
when the mean infectious period is fixed, low growth rate estimates (to match the
lower epidemic peak of the recovery time series) decrease the estimate of the
basic reproduction number.

@@

We assumed that newly infected cases are reported instantaneously when
individuals become infected or when they recover. Although neither of these
assumptions are realistic, they are broadly consistent with the assumptions
of other state-of-art epidemic models. More realistic models may distinguish
reported and unreported cases by adding a new state variable or by modeling
an explicit delay kernel in reporting time.

While we have only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that deaths are recorded as soon as individuals die. This assumption
does not account for the possibility that delays in reporting time of deaths can
change the shape of an epidemic curve: longer delays are expected to decrease
the size of an epidemic peak and delay the timing of the peak (Figure 1). 
These assumptions may have important effects on their conclusions about the underlying 
transmission mechanism.

>>>>

\title{A note on observation processes in epidemic models}
\author{Sang Woo Park and Benjamin M. Bolker}

@@

\section*{Abstract}

@@

Mechanistic analyses of epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an outbreak \citep{breto2009time}. In order to make a 
precise and accurate inference, disease modelers often focus on developing
more realistic process models. For example, the same London measles time 
series from the prevaccination era has been analyzed using multiple models -- 
these models account for time-varying transmission rates \citep{fine1982measles}, 
realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

Despite the amount of effort put into developing \emph{better} process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Most disease models effectively assume that new cases are reported
instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
kennedy2018modeling, pons2018serotype}) or develops symptom (e.g., 
\cite{bhadra2011malaria, king2015avoidable}); 
a few models (e.g.,\cite{breto2009time, he2009plug, lin2016seasonality})
assume that infections are counted upon recovery (because diagnosed cases
are controlled and are effectively no longer infectious).

Here, we use a simple Susceptible-Infected-Removed (SIR) model to study how 
assumptions about the underlying observation processes affect parameter estimates
of the SIR model. We show that making wrong assumptions about the timing of 
incidence reports can lead to biases in parameter estimates and narrow 
confidence intervals.

@@

The Susceptible-Infected-Removed (SIR) model describes how disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate per unit time, $\gamma$ is the recovery rate per unit time, 
and $N = S + I + R$ is the total population size. 
We define incidence at time $t$ as the number of newly infected
individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
the reporting time step. We expect infected individuals to report their infection some time
after their infection; the number of reported cases defines the \emph{observed} incidence.

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection ($i_1(t)$) can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt,
\end{equation}
where $\Delta t$ is the reporting time step.
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model under-reporting using a negative binomial distribution with
means $\rho i_1(t)$ and $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over-dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
likewise, we will refer to epidemic time series generated from these models 
as infection time series and recovery time series.

@@

In order to introduce the problem, we first compare the dynamics of observed 
incidence curves, $i_1(t)$ and $i_2(t)$, (Figure 1). A lag in reporting time delays
the observed epidemic peak timing and reduce the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
would not expect to be able to distinguish between the two reporting processes.

Note that incidence, regardless of when it is measured, is different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the dynamics of observed incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While it is
uncommon, some models do not make a clear distinction between the two (e.g., 
\cite{hooker2010parameterizing}).

\begin{figure}
\includegraphics[width=\textwidth]{../figure/example.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
A deterministic SIR model is simulated using the following parameters: 
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
and $\Delta t = 0.1$ time units.
}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../figure/compare_deterministic.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
We simulate infection time series and recovery time series 100 times using the 
following parameters:  
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
$\rho = 0.5$, $\theta = 10$, and $\Delta t = 0.1$ time units.
For each simulation, we fit the infection model and the recovery model by
(A) allowing for the mean infectious period to be estimated and (B) assuming
that the mean infectious period is known.
Coverage probabilities represent the proportion of confidence intervals
that contain the true value of the basic reproductive number ($\mathcal R_0$).
}
\end{figure}

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in clearer bias and a much lower coverage (Figure 2B); however, 
bias from fitting the wrong model changes direction (e.g., fitting the infection
model to recovery model overestimates the reproduction number instead).

Differences in the direction of the bias can be explained by the estimates of the
exponential growth rate ($r = \beta - \gamma$) and the mean infectious period ($1/\gamma$).
When we fit the recovery model to an infection time series, the model overestimates
the initial growth rate in order to match the higher epidemic peak of the 
infection time series. When the mean infectious period is fixed, higher growth rate
translates to higher basic reproduction number as we see in Figure 2B. When we allow
the mean infectious period to vary, the model also underestimates the mean infectious
period (high $\gamma$), which decreases the overall estimate of the basic reproduction number.
Likewise, fitting the infection model to a recovery time series overestimates the
mean infectious period (low $\gamma$), increasing the estimate of the basic reproduction number;
when the mean infectious period is fixed, low growth rate estimates (to match the
lower epidemic peak of the recovery time series) decrease the overall estimate of the
basic reproduction number.

@@

We assumed that newly infected cases are reported instantaneously when
individuals become infected or when they recover. Although neither of these
assumptions is realistic, they are broadly consistent with the assumptions
of other state-of-art epidemic models (see Introduction). 
More realistic models may distinguish reported and unreported
(identified and unidentified) cases by adding
new state variables \citep{browne2015modeling,webb2015model} 
or by modeling an explicit delay kernel in reporting time 
\citep{harris1990reporting, ferguson2001foot, ster2009epidemiological, 
birrell2011bayesian, funk2018real}.

While we have only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}. 
This assumption does not account for the possibility 
that a delay in reporting time of deaths can
change the shape of an epidemic curve: longer delays are expected to decrease
the size of an epidemic peak and delay the timing of the peak (Figure 1). 
These assumptions may affect their conclusions about the underlying 
transmission mechanisms.
